---
title: "Noncompartmental evaluation of time to steady-state"
author: "Bill Denney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Noncompartmental evaluation of time to steady-state}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PKNCA)
library(knitr)
library(ggplot2)
scale_colour_discrete <- scale_colour_hue
scale_fill_discrete <- scale_fill_hue
scale_colour_ordinal <- scale_colour_hue
scale_fill_ordinal <- scale_fill_hue
```

Time to steady-state (TSS) can be estimated with PKNCA using either a monoexponential
increase toward an asymptote or by a linear regression of the last points.
According to Maganti (2008), the monoexponential method is preferred.

TSS can be estimated using either method using the `pk.tss()` function in PKNCA.

# Example

## Data setup

Illustrating time to steady-state, the example from the
[superposition vignette](v20-superposition.html) will be used.

```{r superposition, error=TRUE}
library(PKNCA)
theoph_corrected <- as.data.frame(datasets::Theoph)
theoph_corrected$conc[theoph_corrected$Time == 0] <- 0
conc_obj <- PKNCAconc(theoph_corrected, conc~Time|Subject)
steady_state <- superposition(conc_obj, dose.times = seq(0, 168 - 12, by=12), tau=168, n.tau=1)
# Add some noise to the data so that it seems more reasonable
steady_state_noise <- steady_state
steady_state_noise$conc <- steady_state_noise$conc*exp(rnorm(nrow(steady_state_noise), mean = 0, sd = 0.1))
```

Examine the data graphically.

```{r superposition-fig}
library(ggplot2)
ggplot(steady_state_noise, aes(x=time, y=conc, groups=Subject)) + geom_line()
```

## Estimate time to Steady State

### Monoexponential

The below code estimates four different types of monoexponential time to
steady-state:

1. tss.monoexponential.population: The population estimate of TSS using a nonlinear mixed effects model (one value for all subjects)
2. tss.monoexponential.popind: The individual estimate from a nonlinear mixed effects model (one value per subject)
3. tss.monoexponential.individual: The individual estimate using a gnls model to estimate each subject separately (one value per subject)
4. tss.monoexponential.single: The mean estiamte of TSS using a nonlinear model

```{r tss-mono}
tss_mono <-
  pk.tss.monoexponential(
    conc = steady_state_noise$conc,
    time = steady_state_noise$time,
    subject = steady_state_noise$Subject,
    time.dosing = seq(0, 168 - 12, by=12)
  )
tss_mono
```

The fraction of steady-state required for time to steady-state can be changed
with the `tss.fraction` argument (see `?pk.tss.monoexponential`).

### Stepwise Linear

The stepwise linear method estimates if the slope of the predose concentrations
is statistically significant starting from the last measurement and moving
backward in time.  It has bias in that more individuals will move the time to
steady-state to a late time point.

```{r tss-step}
tss_step <-
  pk.tss.stepwise.linear(
    conc = steady_state_noise$conc,
    time = steady_state_noise$time,
    subject = steady_state_noise$Subject,
    time.dosing = seq(0, 168 - 12, by=12)
  )
tss_step
```

# References

1. Maganti, L., Panebianco, D.L. & Maes, A.L. Evaluation of Methods for Estimating Time to Steady State with Examples from Phase 1 Studies. AAPS J 10, 141–147 (2008). https://doi.org/10.1208/s12248-008-9014-y
